---
title: "Tidy Nutri Data"
author: "Saghir & Andreia (ilustat)"
date: "`r Sys.time()`"
output:
  html_document: default
  pdf_document: default
---
<!-- date: "`r format(Sys.time(), "%d %B %Y (%X)")`" -->

***

**Start: `r format(Sys.time(), "%d %B %Y (%X)")`**

***


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(stringr)
library(emojifont)
```

## Importing Nutri Data

Food composition data is publically available and provided by **Instituto Nacional de Saúde Dr. Ricardo Jorge** (portuguese "National Institute of Health" - INSA): http://portfir.insa.pt/foodcomp/introduction 
*It should not be used with any commercial intent.*

To download the data you can follow these steps:
Access to "Composição dos alimentos" tab, then select 'Pesquisa de Alimentos' >> 'Download da TCA'

Save it in the folder `data` under the same working directory of the `nutrient_pt` project and read it as follows:

```{r importNutriData}
nutri_orig <- read_xlsx("data/insa_tca.xlsx") 
#glimpse(nutri_orig)
```

## Problem to make us cry `r emoji("cry")`

*Problem:* Variable names of the the original data and the risk of incorrectly tidying the data.

When we look at the `nutri_long` dataset we see that we could lose the ordering
of the `keyVars`. We need to protect again this causing a problem during the data
tidying. The problem stems from the original data, looking at the names of `nutri_orig`:

```{r namesOrig}
names(nutri_orig)[1:12]
```

Variable `Energia [kcal] (ENERCC)` contains the value for nutrient "Energia [kcal] (ENERCC)", 
`X__2` the associated unit, and `X__3` the associated quantity. It is the same for variable `Lípidos (FAT)` 
contains the value for nutrient "Lípidos (FAT)", `X__6` the associated unit, and `X__7` the associated quantity. This is the same pattern for all the nutrients as it originates from the original Excel
spreadsheet where each nutrients has 3 columns but one *header* in a merged cell above the 3 columns.

Using the column names of the the original data we will create an ordering variable.
First we will remove the the first three variables as they will remain _constants_ as columns.

The we will create a grouping variable for the nutrients value, unit and quantity called `grpNtr`
and `grpType`.

```{r ordNames}
ordNames <- as_tibble(list(varName = names(nutri_orig)[-c(1, 2, 3)])) %>%
  mutate(grpNtr = ceiling(row_number()/3)) %>% 
  mutate(typeVars = parse_number(varName) %% 2) %>% 
  mutate(xVars = str_detect(varName, "^X__")) %>% 
  mutate(grpType = if_else(typeVars == 1, "Quantity", "Unit")) %>% 
  mutate(grpType = if_else(xVars == FALSE, "Nutrient", grpType)) %>% 
  select(-xVars, -typeVars)
  
head(ordNames, n=12)
```

## Tidying Data Step 1

From the `nutri_orig` data we will create a long format 

Now we will merge the group ordering to the `nutri_long` dataset

```{r}
nutri_long <- nutri_orig %>% 
  rename(foodID = X__1, 
         foodItem = `Nome do alimento`, 
         foodGroup = Grupo) %>% 
  gather(key = "keyVars", value = "keyVals", -starts_with("food")) %>% 
  select(foodID, foodGroup, foodItem, keyVars, keyVals) %>% 
  left_join(ordNames, by= c("keyVars" = "varName")) 

glimpse(nutri_long)
```


## Tidying Nutri Data

In the `nutri_long` dataset we have `r format(nrow(nutri_long))` observations. Our final 
tidy dataset, which we call `nutri_tidy` should contain one third of these values (i.e., 
`r format(nrow(nutri_long)/3)`).

```{r tidyNutriData}
# Prepare the Units and Quantity so that they can be merged to the Nutrient data.
nUnit <- nutri_long %>% 
  filter(grpType == "Unit") %>% 
  select(foodID, grpNtr, Unit = keyVals) 

head(nUnit)

nQty <-  nutri_long %>% 
  filter(grpType == "Quantity") %>% 
  select(foodID, grpNtr, Quantity = keyVals) 

head(nQty)

# Prepare the base for the Nutrition data by keeping only the nutrient and values
nutri_tidy <- nutri_long %>% 
  filter(grpType == "Nutrient") %>% 
  rename(Value = keyVals,
         Nutrient = keyVars) %>% 
  left_join(nUnit, by = c("foodID", "grpNtr")) %>% 
  left_join(nQty, by = c("foodID", "grpNtr")) %>% 
  # select(-grpNtr, -grpType) %>% 
  mutate(NutrientCode = gsub(".*\\((.*)\\).*", "\\1", Nutrient),
         NutrientID = group_indices(. , Nutrient)) %>% 
  mutate(Value = as.numeric(Value)) %>% 
  select(foodID, foodGroup, foodItem, NutrientID, Nutrient, NutrientCode, Value, Unit, Quantity) %>%
  arrange(foodID, Nutrient)

  glimpse(nutri_tidy)
```


## Cleaning Data

Data is tidy but not totally cleaned yet
```{r}
nutri_tidy %>% arrange(foodGroup, foodItem) %>% select(foodItem)
```

We use `stringr` to clean up the values from `foodItem` where we can still find quotation marks ("") 
```{r}
nutri_clean <- nutri_tidy %>%
        # clean observations
        mutate(foodItem = str_replace_all(
                foodItem, "\\p{quotation mark}", ""),
               Nutrient = str_replace_all(
                       Nutrient, "\\+", ""),
               # remove square brackets to facilitate filtering detection later
               Nutrient = str_replace_all(
                       Nutrient, "\\s[\\[\\]]", "_"),
               Nutrient = str_replace_all(
                       Nutrient, "[\\]]", "")
               ) %>% 
        arrange(foodGroup, foodItem)
# DT::datatable(nutri_clean)
```

Now when we search for "Donut" on the search window we no longer see it with quotation marks.

Save the cleaned dataset to use it later in the Shiny app
```{r, eval = FALSE}
#save(nutri_clean, file = "data/nutri_clean.RData")
```

```{r}
#load("nutri_clean.RData")
```

Get data into wide format for food items
```{r}
nutri_wide <- nutri_clean %>%
        select(foodID, foodItem, NutrientID, Nutrient, Value, Unit, -Quantity) %>%
        group_by(foodItem) %>%
        select(-foodID) %>%
        spread(foodItem, Value) 

head(nutri_wide)
```

Save the wide version to use it later in the Shiny app

```{r, eval = FALSE}
#save(nutri_wide, file = "data/nutri_wide.RData")
```

Selecting/Matching specific variables 

```{r}
x <- c("Abacate", "Abóbora", "Açorda")

nutri_wide %>% select(Nutrient, str_subset(names(.), x)) %>% head()
```

**Unite variables unit with Food to enable later spread of nutrient values into individual variables**

```{r}
nutri_new <- nutri_clean %>%
        mutate(Quantity_unit = str_detect(Quantity, "g"),
               Quantity_unit = ifelse(Quantity_unit == TRUE, "g", "mL"),
               Quantity = str_replace_all(Quantity, "[^[\\d]]", ""),
               Value = sprintf("%6.f", Value),
               Value = as.numeric(Value),
               Quantity = as.numeric(Quantity)) %>%
        group_by(foodItem) %>%
        unite(Food, foodItem, Quantity_unit, sep = " (") %>%
        mutate(Food = str_c(Food, ")")) %>%
        select(FoodID = foodID, Food, FoodGroup = foodGroup, Quantity, NutrientID, Nutrient, Unit, Value) 
head(nutri_new)
        
#save(nutri_new, file = "data/nutri_new.RData", envir = .GlobalEnv)

```

Example of using `nutri_new` in Shiny for user input represented by `y`
```{r}
y <- "Energia_kcal (ENERCC)"
nutri_comp <- nutri_new %>%
                        select(Food, Quantity, Nutrient, Unit, Value) %>%
                        unite(Nutrient, Nutrient, Unit, sep = " (") %>%
                        mutate(Nutrient = str_c(Nutrient, ")")) %>%
                        spread(Nutrient, Value) %>%
                        select(Food, Quantity,
                               str_subset(names(.),
                                          str_c(str_match(y,
                                                          "^[\\w-\\w+\\s]+"),
                                                          collapse = "|")))
head(nutri_comp)
```


***

**End: `r format(Sys.time(), "%d %B %Y (%X)")`**

***

