## Where are the calories coming from?

If we convert the macronutrients weight unit (grams) into calories we can filter foods for highest/lowest percentage of calories from each macronutrient (protein, lipids, carbohydrates).

**Load required libraries**
```{r, warning=FALSE, message = FALSE}
library(tidyverse)
library(stringr)
library(here) # to load data stored in a different level directory
library(DT)
```

**Load required datasets (pre-processed tidy data)**
```{r}
load(here("data/food_wide.RData"))
load(here("data/nutri_choice.RData"))
```


### Which food items have more calories from fat/sugar/protein

Select only macronutrients 
```{r}
nutri_macro <-  food_wide %>% 
        select(FoodID, Food, str_subset(names(.), "ENERCC|\\(CHO\\)|PROT|\\(FAT\\)"))

head(nutri_macro)

```


#### Energy Conversions

Conversions will be made according to [FAO agreement on Food energy conversion factors](http://www.fao.org/docrep/006/Y5022E/y5022e05.htm#TopOfPage)

**Prepare a table with the energy conversion factors for each nutrient**
```{r}
names <- names(nutri_macro)[3:6]
names
nutri_conv <- tibble(Nutrient = names, Calories_Factor = c(1, 4, 9, 4))
nutri_conv
```

```{r}
save(nutri_conv, file = here("data/nutri_conv.RData"))
```


**Join energy conversion factors into the whole dataset** 
```{r}
nutri_join <- nutri_choice %>%
        filter(Nutrient %in% names) %>%
        left_join(nutri_conv, by = 'Nutrient')
knitr::kable(head(nutri_join))
```

**Calculate energy values for each nutrient**
```{r}
nutri_cal <- nutri_join %>%
        mutate(Value_cal = Value * Calories_Factor) 
DT::datatable(nutri_cal)
```


**Get only the value of total energy (quilocaloria) to determine energy percentages for each macronutrient**
```{r}
knitr::kable(head(nutri_macro))

energy <- nutri_macro %>% 
        select(FoodID, Food, str_subset(names(.), "ENERCC"))
```

**Join the energy total values with previous macronutrients dataset to determine percentages**
```{r}
nutri_perc <- nutri_cal %>% 
        left_join(energy, by = c('FoodID', 'Food')) %>%
        mutate(Value_perc = sprintf("%6.1f", (Value_cal * 100)/ `Energia_kcal (ENERCC) (quilocaloria)`))
DT::datatable(nutri_perc)
```

